Stability Analysis of Recurrent Neural Networks \\ using Ordinary Differential Equations

### 2020-09-23 p5/stability_rnn.tex
Final draft

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol. Whenver
a symbol stands for a vector, it denotes a column vector. $\mathbb{R}$ is 
the set of real numbers. $\mathbb{Z}$ is the set of integers. If $z$ is
a complex number then $\Re(z)$ denotes its real part and $\Im(z)$ its
imaginary part.

\section{Ordinary Differential Equations of the first order}\label{s3}
An ordinary differential equation, abbreviated as ODE, of the first order 
is an equation of the form
\begin{equation}\label{s3e1}
\frac{dy}{dx} = f(x, y),
\end{equation}
where $x \in \mathbb{R}$, $y$ is the real-valued unknown function, $f: 
\mathbb{R}^2 \mapsto \mathbb{R}$. Note that $f$ depends on both $x$ and 
$y$. It called a first order ODE because the highest order derivative
of the unknown function $y$ with respect to $x$ is of the first order.
This equation is usually accompanied with the initial value of $y$ at a 
certain $x$, say $x = x_0$. Equation \eqref{s3e1} and $y(x_0) = y_0$ is
called an \emph{inital value problem}.

@@

\section{Runge-Kutta methods}\label{s6}
Runge-Kutta methods are a collection of numerical algorithms developed by
the German mathematicians Carl Runge and Martin Kutta in the late 
nineteenth century\cite{lambert1991numerical}. We will describe the most
commonly used among them, called the `RK4' method. If $y: \mathbb{R}
\mapsto \mathbb{R}$ is the unknown function that obeys the ODE
\begin{equation}\label{s6e1}
\frac{dy}{dt} = f(t, y), y(t_0) = y_0
\end{equation}
then one can get the value oy $y$ at $t_n + h$ using the formula
\begin{equation}\label{s6e2}
y(t_n + h) = y(t_n) + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4).
\end{equation}
The constants $k_1, \ldots, k_4$ are given by
\begin{eqnarray}
k_1 &=& f(t_n, y(t_n)) \label{s6e3} \\
k_2 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_1}{2}\right) 
 \label{s6e4} \\
k_3 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_2}{2}\right) 
 \label{s6e5} \\
k_4 &=& f(t_n + h, y_n + hk_3). \label{s6e6}
\end{eqnarray}
Let us study these equations carefully. The value of $y$ at $t_n + h$ 
depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Each of
these is the value of the derivative of $y$ at different points. Of these,
$k_1$ depends on $y(t_n)$ and $t_n$, $k_2$ on $y(t_n)$, $t_n$ and $k_1$,
$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. If
one were to implement this algorithm, every iteration of it would consist
of using the output of the previous iteration and a calculation of four
constants, one after another, to get the value of the current iteration.
This is indeed like a recurrent neural network with four hidden layers.

@@

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because the
hidden layers are not connected to each other. The Burrage-Butcher tensor 
is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semidefinite 
unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
observation also corroborated by elsewhere\cite{kanai2017preventing}.

@@

The stability of Runge-Kutta methods refers to their ability to generate
contractive solutions in an interval. If a method is not stable it does not
imply that it cannot generate \emph{any} contractive pair of solutions. All
it means is that \emph{not all} pairs are contractive in the sense of 
definition \ref{s9d2}. Existence of one, or a few, convergent solutions to 
the algorithm is not a contradiction to it being unstable. An instability 
means that \emph{not all} solutions are stable. On the other hand stability 
means that \emph{all} solutions are stable. An existence of a few coverging 
solutions does not guarantee the stability of the algorithm. However, an 
existence of a single unstable solution renders an algorithm unstable.

@@

\subsection{Example 1}
We choose the following values of parameters, (first row of table I in 
\cite{jordan2019gated}),
\begin{eqnarray}
U_h &=& 3I_2 \label{s11e12} \\
U_r &=& 0 \label{s11e13} \\
U_z &=& 0 \label{s11e14} \\
b_h &=& 0 \label{s11e15} \\
b_r &=& 0 \label{s11e16} \\
b_z &=& 0 \label{s11e17} 
\end{eqnarray}
For these values $z(t) = r(t) = 1/2(1, 1)^T$ and
\begin{equation}\label{s11e18}
\dot{h}(t) = \frac{1}{2}\left(-h(t) + \tanh\left(\frac{3}{2}h(t)\right)\right)
\end{equation}
The stable fixed points of this equation are shown in figure \ref{f1}.
\begin{figure}[h]
\caption{Stable points}\label{f1}
\includegraphics[scale=0.5]{Ex1_Stable_points.png}
\centering
\end{figure}
The coloured dots indicate the start of the trajectory while the blue square
indicates its end. Trajectories seem to converge on the stable points. The
phase portrait also indicates that the arrows point toward the stable points.

This equation also has an unable point at the origin. The phase portrait has
arrows diverging from the origin. The trajectories follow these directions.
\begin{figure}[h]
\caption{Unstable points}\label{f2}
\includegraphics[scale=0.5]{Ex1_Unstable_point.png}
\centering
\end{figure}
We will now examine the behaviour of this unstable point for a collection of
five such GRUs starting their trajectory in the vicinity of the origin.
In each of these figures the trajectory of the GRUs with attention layer
is shown as a dashed line ending with a red square. The trajectories of the
individual GRUs are shown as solid lines ending at the blue square. The
initial conditions of the GRUs were chosen randomly from a set of $16$ points
on a square centred at the origin and of side $0.1$. Figure \ref{f3} shows
the first set of $6$ samples and figure \ref{f4} shows the remaining $6$
samples.

@@

\begin{figure}[h]
\caption{Unstable point with attention layer - set 1}\label{f3}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_0}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_1}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_2}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_3}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_4}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_5}
\end{array}$
\end{center}
\end{figure}

\begin{figure}[h]
\caption{Unstable point with attention layer - set 2}\label{f4}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_6}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_7}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_8}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_10}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\end{figure}

\subsection{Example 2}
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}

>>>>

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol. Whenever
a symbol stands for a vector, it denotes a column vector. $\mathbb{R}$ is 
the set of real numbers. $\mathbb{Z}$ is the set of integers. If $z$ is
a complex number then $\Re(z)$ denotes its real part and $\Im(z)$ its
imaginary part.

\section{Ordinary Differential Equations of the first order}\label{s3}
An ordinary differential equation, abbreviated as ODE, of the first order 
is an equation of the form
\begin{equation}\label{s3e1}
\frac{dy}{dx} = f(x, y),
\end{equation}
where $x \in \mathbb{R}$, $y$ is the real-valued unknown function, $f: 
\mathbb{R}^2 \mapsto \mathbb{R}$. Note that $f$ depends on both $x$ and 
$y$. It called a first order ODE because the highest order derivative
of the unknown function $y$ with respect to $x$ is of the first order.
This equation is usually accompanied with the initial value of $y$ at a 
certain $x$, say $x = x_0$. Equation \eqref{s3e1} and $y(x_0) = y_0$ is
called an \emph{initial value problem}.

@@

\section{Runge-Kutta methods}\label{s6}
Runge-Kutta methods are a collection of numerical algorithms developed by
the German mathematicians Carl Runge and Martin Kutta in the late 
nineteenth century\cite{lambert1991numerical}. We will describe the most
commonly used among them, called the `RK4' method. If $y: \mathbb{R}
\mapsto \mathbb{R}$ is the unknown function that obeys the ODE
\begin{equation}\label{s6e1}
\frac{dy}{dt} = f(t, y), y(t_0) = y_0
\end{equation}
then one can get the value of $y$ at $t_n + h$ using the formula
\begin{equation}\label{s6e2}
y(t_n + h) = y(t_n) + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4).
\end{equation}
The constants $k_1, \ldots, k_4$ are given by
\begin{eqnarray}
k_1 &=& f(t_n, y(t_n)) \label{s6e3} \\
k_2 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_1}{2}\right) 
 \label{s6e4} \\
k_3 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_2}{2}\right) 
 \label{s6e5} \\
k_4 &=& f(t_n + h, y_n + hk_3). \label{s6e6}
\end{eqnarray}
Let us study these equations carefully. The value of $y$ at $t_n + h$ 
depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Each of
these is the value of the derivative of $y$ at different points. Of these,
$k_1$ depends on $y(t_n)$ and $t_n$, $k_2$ on $y(t_n)$, $t_n$ and $k_1$,
$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. If
one were to implement this algorithm, every iteration of it would consist
of using the output of the previous iteration and a calculation of four
constants, one after another, to get the value of the current iteration.
This is indeed like a recurrent neural network with four hidden layers.

@@

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because the
hidden layers are not connected to each other. The Burrage-Butcher tensor 
is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semi-definite 
unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
observation also corroborated by elsewhere\cite{kanai2017preventing}.

@@

The stability of Runge-Kutta methods refers to their ability to generate
contractive solutions in an interval. If a method is not stable it does not
imply that it cannot generate \emph{any} contractive pair of solutions. All
it means is that \emph{not all} pairs are contractive in the sense of 
definition \ref{s9d2}. Existence of one, or a few, convergent solutions to 
the algorithm is not a contradiction to it being unstable. An instability 
means that \emph{not all} solutions are stable. On the other hand stability 
means that \emph{all} solutions are stable. An existence of a few converging 
solutions does not guarantee the stability of the algorithm. However, an 
existence of a single unstable solution renders an algorithm unstable.

@@

\subsection{Example 1}
We choose the following values of parameters, (first row of table I in 
\cite{jordan2019gated}),
\begin{eqnarray}
U_h &=& 3I_2 \label{s11e12} \\
U_r &=& 0 \label{s11e13} \\
U_z &=& 0 \label{s11e14} \\
b_h &=& 0 \label{s11e15} \\
b_r &=& 0 \label{s11e16} \\
b_z &=& 0 \label{s11e17} 
\end{eqnarray}
For these values $z(t) = r(t) = 1/2(1, 1)^T$ and
\begin{equation}\label{s11e18}
\dot{h}(t) = \frac{1}{2}\left(-h(t) + \tanh\left(\frac{3}{2}h(t)\right)\right)
\end{equation}
The stable fixed points of this equation are shown in figure \ref{f1}.
\begin{figure}[!htbp]
\caption{Stable points}\label{f1}
\includegraphics[scale=0.5]{Ex1_Stable_points.png}
\centering
\end{figure}
The coloured dots indicate the start of the trajectory while the blue square
indicates its end. Trajectories seem to converge on the stable points. The
phase portrait also indicates that the arrows point toward the stable points.

This equation also has an unstable point at the origin. The phase portrait 
has arrows diverging from the origin. The trajectories follow these 
directions.
\begin{figure}[!htbp]
\caption{Unstable points}\label{f2}
\includegraphics[scale=0.5]{Ex1_Unstable_point.png}
\centering
\end{figure}
We will now examine the behaviour of this unstable point for a collection of
five such GRUs starting their trajectory in the vicinity of the origin.
In each of these figures the trajectory of the GRUs with attention layer
is shown as a dashed line ending with a red square. The attention layer is 
connected in the architecture proposed by Ran and others, described previously
in the first subsection of section \ref{s10}. Equations \eqref{s10e1} to 
\eqref{s10e4} were used to generate the final output. We used a randomly 
generated weight matrix $W_H$. Since $W_v$ and $A$ are fixed we chose their
product to be a fixed, but randomly generated, vector in $\mathbb{R}^2$. The
weight matrix $W$, whose rows are used in equation \eqref{s10e3} too was 
selected randomly.

The trajectories of the
individual GRUs are shown as solid lines ending at the blue square. The
initial conditions of the GRUs were chosen randomly from a set of $16$ points
on a square centred at the origin and of side $0.1$. Figure \ref{f3} shows
the first set of $6$ samples and figure \ref{f4} shows the remaining $6$
samples.

@@

\begin{figure}[!htbp]
\caption{Unstable point with attention layer - set 1}\label{f3}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_0}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_1}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_2}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_3}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_4}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_5}
\end{array}$
\end{center}
\end{figure}

\begin{figure}[!htbp]
\caption{Unstable point with attention layer - set 2}\label{f4}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_6}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_7}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_8}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_10}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\end{figure}

\subsection{Example 2}
We now choose the parameters in the last row in table I of 
\cite{jordan2019gated}. Thus,
\begin{eqnarray}
U_h &=& \begin{pmatrix}2.9665 & -0.4771 \\ 0.4771 & 2.9665\end{pmatrix}
         \label{s11e19} \\
U_r &=& 0 \label{s11e20} \\
U_z &=& 0 \label{s11e21} \\
b_h &=& 0 \label{s11e22} \\
b_r &=& 0 \label{s11e23} \\
b_z &=& 0 \label{s11e24} 
\end{eqnarray}
The ODE for these parameters is
\begin{equation}\label{s11e25}
\dot{h}(t) = \frac{1}{2}\left(-h(t) + \tanh\left(\frac{1}{2} 
\begin{pmatrix}2.9665h_x-0.4471h_y \\0.4471h_x+2.9665h_y\end{pmatrix}\right)
\right)
\end{equation}
The phase portrait of this equation is shown in figure \ref{f5}.
\begin{figure}[!htbp]
\caption{Unstable points}\label{f5}
\includegraphics[scale=0.5]{Ex2_stable_points.png}
\centering
\end{figure}
No fixed points are visible in the domain shown in the figure although the
trajectories seem to be converging asymptotically toward a few lines. The
phase portrait nevertheless shows the possibility of the origin being an
unstable point. The derivative of the function seems to be diverging from the
origin in all directions. We examine the nature of this differential equation
in the neighbourhood of the origin in figure \ref{f6}.
\begin{figure}[!htbp]
\caption{Unstable points}\label{f6}
\includegraphics[scale=0.5]{Ex2_Unstable_point.png}
\centering
\end{figure}
The trajectories are indeed diverging away from the origin. Trajectories
starting from initial conditions close to each other end up going very far
from each other. This is indeed a signature of an instability.

We repeat the experiment we did in the previous example in which we attach
an attention layer to five GRUs with different initial conditions, all in the
neighbourhood of the origin. Once again we observe that the trajectory of the
GRU's with attention layer does not reach a point farther than the trajectories
of the individual GRUs.

\begin{figure}[!htbp]
\caption{Unstable point with attention layer - set 1}\label{f7}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_0}&
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_1}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_2}&
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_3}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_4}&
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_5}
\end{array}$
\end{center}
\end{figure}

\begin{figure}[!htbp]
\caption{Unstable point with attention layer - set 2}\label{f8}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_6}&
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_7}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_8}&
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_10}&
\includegraphics[scale=0.4]{Ex2_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\end{figure}

\subsection{Remarks on the experiments}
Recall the idea of stability of an ODE. If $y_1$ and $y_2$ are two solutions
of the ODE with initial conditions close to each other then if $y_1$ and $y_2$
remain close to each other for ever then we say that the ODE is stable in a
neighbourhood of the two initial conditions. On the other hand, if $y_1$ and 
$y_2$ go very far from each other then it means that a slight change in the
initial conditions pushes the solutions very far from each other. It is clear
from figures \ref{f3}, \ref{f4}, \ref{f7} and \ref{f8} that the individual 
solutions of the ODEs tend to diverge from each other. Coupling a few ODEs 
(GRUs) to an attention layer does not eliminate this behaviour. The dashed 
lines indeed go in a different direction in every figure. But their final 
destinations, marked by a red square, are closer to each other than the
final destinations of the bare GRUs.

\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}
